From Reversible Programs to \\ Univalent Universes and Back

### 2021-06-29 Pi+/paper/ufin.tex
Minor: fix build
Revise section 4.2

We have $\Fin[3] & = \Set{0,1,2} \eqv \unit \sqcup (\unit \sqcup \unit)$ which fixes a particular enumeration of the
elements. Suppose we have a set $X = (\unit \sqcup \unit) \sqcup \unit$, it has the same cardinality as $\Fin[3]$, so it
is represented by the same 0-cell. But, $X$ can be made equivalent to $\Fin[3]$ in many different ways since there are
many bijections between them. One bijection is
$\Set{\inl(\inl(\ttt)) \mapsto 0, \inl(\inr(\ttt)) \mapsto 1, \inr(\ttt) \mapsto 2}$ which can be written in two
different ways by composing more primitive operations, $f_{1} = \assocrp$, or
$f_{2} = \swapp \compc \assoclp \compc \swapp$. Another bijection is
$\Set{\inl(\inl(\ttt)) \mapsto 1, \inl(\inr(\ttt)) \mapsto 2, \inr(\ttt) \mapsto 0}$ which is given by $f_{3} = \ldots$.
Since $f_{1}$ and $f_{2}$ produce the same enumeration of the elements of $X$, they are identified by a homotopy $h$
which is encoded in the 2-cell between them.

@@

In order to define the semantics, we seek a mathematical structure satisfying the following properties: (i) it contains structures corresponding to all the finite types and nothing but the finite types, and (ii) it is robust enough to ensure that equivalent encodings of finite types are identified. \note{Is this the correct informal explanation of Ufin?}

@@

We work in Homotopy Type Theory, in particular, intensional Martin-L\"{o}f Type Theory, with a univalent universe, and
Higher Inductive Types (HITs) for propositional truncations and set quotients. We recall a few basic facts to
familiarise the reader with the notation we use. For more details about basics of HoTT, we refer the reader to the HoTT
book~\cite{univalentfoundationsprogramHomotopyTypeTheory2013}.

\note{It is important that we work in HoTT using it as a metatheory, we care about proof relevance, constructivity, and
  working with higher-dimensional algebraic structures, such as groupoids. Our results are also formalised using general
  tools and principles from HoTT, and computer checked using a proof assistant, using Agda and the HoTT-Agda library.}

@@

In HoTT, the intensional identity type is the type of paths between two terms of the same type. Given two terms $x:A$
and $y:A$, we write $x \id_{A} y$, or simply $x \id y$, for the equality type between them. In book HoTT, the identity
type is generated by reflexivity $\refl_{x}$, and the eliminator for the identity type is given by path induction or the
$J$-rule. The identity type equips each type with the structure of an infinity groupoid, or a homotopy type.

\todo{Say why it is a groupoid by listing some groupoid laws.}

Functions between types are functors between groupoids. Given a function $f : A \to B$, the functorial action is given
by

@@

Type families are functions from a type to the universe, that is, an indexed family of groupoids. The $\term{transport}$
operation gives the functorial action of paths in the indexing type, which is defined by path induction. If
$P : A \to \UU$ is a type family, then for a path $x \id_{A} y$, we have

@@

\todo{Why is the topological viewpoint important, or saying fibrations? Explain the intuition using classifying spaces,
  maybe draw some diagrams.}

From the topological viewpoint, a type family can also be seen as a fibration. For a type family $P : A \to \UU$ and a
point $x : A$, $P(x)$ gives the fiber over $a$. For a path $p : x \id_{A} y$, $\transport{P}{p}$ gives the path lifting
operation. The total space is given by $\dsum{x:A}{P(x)}$ and the first projection ${\pi_1 : \dsum{x:A}{P(x)} \to A}$ to
the base space is the fibration. The lifting operation lifts paths in the base space to paths in the total space. If
$p : x \id_{A} y$ is a path in the base space, and $u : P(x)$, we have

@@

Further, using the groupoid structure of $A$, we can show that transport lifts paths to equivalences, we define

@@

\todo{Explain motivation.}

>>>>

We have $\Fin[3] = \Set{0,1,2} \eqv \unit \sqcup (\unit \sqcup \unit)$ which fixes a particular enumeration of the
elements. Suppose we have a set $X = (\unit \sqcup \unit) \sqcup \unit$, it has the same cardinality as $\Fin[3]$, so it
is represented by the same 0-cell. But, $X$ can be made equivalent to $\Fin[3]$ in many different ways since there are
many bijections between them. One bijection is
$\Set{\inl(\inl(\ttt)) \mapsto 0, \inl(\inr(\ttt)) \mapsto 1, \inr(\ttt) \mapsto 2}$ which can be written in two
different ways by composing more primitive operations, $f_{1} = \assocrp$, or
$f_{2} = \swapp \compc \assoclp \compc \swapp$. Another bijection is
$\Set{\inl(\inl(\ttt)) \mapsto 1, \inl(\inr(\ttt)) \mapsto 2, \inr(\ttt) \mapsto 0}$ which is given by $f_{3} = \ldots$.
Since $f_{1}$ and $f_{2}$ produce the same enumeration of the elements of $X$, they are identified by a homotopy $h$
which is encoded in the 2-cell between them.

@@

In order to define the semantics, we seek a mathematical structure satisfying the following properties: (i) it contains structures corresponding to all the finite types and nothing but the finite types, and (ii) it is robust enough to ensure that equivalent encodings of finite types are identified.

@@

We work in Homotopy Type Theory (HoTT), and use the language of the HoTT
book~\cite{univalentfoundationsprogramHomotopyTypeTheory2013}, that is, we use intensional Martin-L\"{o}f Type Theory,
with a univalent universe $\UU$, and we use Higher Inductive Types (HITs) for propositional truncations and set
quotients. All arguments will hold in a cubical type theory as well, such
as~\cite*{cohenCubicalTypeTheory2018,angiuliComputationalSemanticsCartesianCubical2019}. We recall a few basic facts to
familiarise the reader with the notation we use. For more details about basics of HoTT, we refer the reader to the HoTT
book~\cite{univalentfoundationsprogramHomotopyTypeTheory2013}.

Our denotational semantics is given using groupoids, and HoTT provides a rich internal language for describing (weak
$\infty$-)groupoids. By formalising our results using general tools and principles from HoTT, we're able to use a proof
assistant to computer-check our proofs. Our results are formalised in the Agda proof assistant using the HoTT-Agda
library, though it could also be formalised using Cubical Agda~\cite{vezzosiCubicalAgdaDependently2019}. \review{Using
  the formalisation of the semantics, we are also able to extract a computable procedure for performing normalisation of
  circuits.}

@@

In HoTT, the identity type or the equality type is the type of paths or equalities between two terms of the same type.
Given two terms $x:A$ and $y:A$, we write $x \id_{A} y$, or simply $x \id y$, for the equality type between them. In
book HoTT, the identity type is generated by reflexivity $\refl_{x} : x \id_{A} x$, and the eliminator for the identity
type is given by path induction or the $J$-rule. The identity type equips each type with the structure of a weak
$\infty$-groupoid, or a homotopy type. For example, given an arbitrary type (or groupoid) $A$, we list some groupoid
laws that are provable using path induction.

\begin{gather*}
  \begin{aligned}
    \term{sym} & : (x \id_{A} y) \to (y \id_{A} x) \\
    \term{\comp} & : (x \id_{A} y) \to (y \id_{A} z) \to (x \id_{A} z)
  \end{aligned}
  \qquad
  \begin{aligned}
    \term{assoc} & : (p : x \id_{A} y)  (q : y \id_{A} z) (r : z \id_{A} w) \\
    & \to (p \comp q) \comp r \id p \comp (q \comp r) \\
    \term{invr} & : (p : x \id_{A} y) \to p \comp \term{sym}(p) \id \refl_{x}
  \end{aligned}
\end{gather*}

Functions between types are functors between groupoids. Given a function $f : A \to B$, the functorial action
$\term{ap}_{f}(\blank)$ is given by path induction.

@@

Type families are functions from a type to the universe, such as $A \to \UU$, which is an $A$-indexed family of
groupoids. The $\term{transport}$ operation gives the functorial action of paths in the indexing type, which is also
defined by path induction. If $P : A \to \UU$ is a type family, then for a path $x \id_{A} y$, we have

@@

\todo{Diagram.}

A type family can also be seen as a fibration, from the topological viewpoint. For a type family $P : A \to \UU$ and a
point $x : A$, the type $P(x)$ gives the fiber over $x$. The type $\dsum{x:A}{P(x)}$ is the collection of all the fibers
and is called the total space of $P$. The first projection ${\pi_1 : \dsum{x:A}{P(x)} \to A}$ from the total space to
the base space $A$ is then a fibration. The lifting operation lifts paths in the base space to paths in the total space.
Given a path $p : x \id_{A} y$ in the base space, and $u : P(x)$ a point in the fiber over $x$, we have

@@

Further, using the groupoid structure of $A$, we can show that $\term{transport}$ lifts paths to equivalences, we define

@@

\todo{Diagram.}
